--- 
title: "Incorporating Spatial Analysis into Agricultural Field Experiments."
author: "Julia Piaskowski"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This includes instructions for diagnosing spatial variation in agricultural field trials and incorporating spatial variation into analysis of the trial using freely available tools in R."
---

# Preface 

## Tutorial goals

To help people conducting planned agricultural field trials to understand and  incorporate spatial variation routinely into analysis of field trials.  

current resources are focused on geospatia appications - LANDSAT data across large area, point information on soils and geology, and other information that typiically requiresa a moderate to deep understanding of obht GIS tools and spatial analytica techniques. 

Furthermore, there is not a comprehensive one-stop-shop for spatial analytical techniques for fied experiments that is also freely available. 





## Prerequisites

In order to run the scripts in this demonstration, you will to download R, avaiable free through the [Comprehensive R Archive Network](). While this is sufficient for running R scripts, You may also find it hepful to use RStudio, which provides a nice graphical user interface for R. RStudio can be downloaded [here]().

If you already have R installed, please make sure you have version 3.5.0 or newer. 

This demonstration is not intended to provide instructions on general R usage. However, There are numerous web resources for learning the Basics of R:

* 
* 
* 
*

## Packages we will use 


## Acknowledgements

## License

 



#### OLD STUFF

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
